---
output: github_document
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# VIUSr

<!-- badges: start -->
[![R-CMD-check](https://github.com/jashonnew/VIUSr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jashonnew/VIUSr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## Overview

**`VIUSr`** is an R package developed to facilitate the cleaning, exploration, and visualization of data from the 2021 Vehicle Inventory and Use Survey (VIUS) conducted across the United States. This package provides a suite of tools to streamline the workflow from raw data to actionable insights, supporting users in extracting meaningful patterns at both national and state levels.

## Key Features

- **Automated Data Cleaning**: Efficiently prepares raw VIUS 2021 data into a tidy, analysis-ready format with minimal user intervention. Cleaning functions added include `convert_names`, `drop_bad_values`, and `charToNum`.
- **State-Level Exploration**: Generate focused summaries and visualizations for individual U.S. states to understand regional vehicle usage patterns.
- **Cross-State Comparison**: Compare metrics across states to identify trends and outliers in vehicle inventory and usage.
- **National Overview**: Access high-level summaries and visual representations of key statistics across the entire U.S.
- **Built-In Visualization Tools**: Create interactive and publication-ready plots, including maps and charts, with intuitive functions designed for ease of use.

`VIUSr` is ideal for transportation analysts, researchers, policy makers, and data professionals seeking to work with VIUS 2021 data in a reproducible and interpretable manner.

While many columns from the VIUS dataset are included in our cleaned version of the 2021 data within the package, using a future year's data with the package or using any of the columns that we chose not to include will require a local download of the dataset upon which you can then apply our cleaning functions and subsequently use our graphing functions. 


## Installation

You can install the development version of VIUSr from [jashonnew/VIUSr](https://github.com/jashonnew/VIUSr) with:

``` r
install.packages("jashonnew/VIUSr")
```

## Data Cleaning

This package contains a cleaned version of the 2021 VIUS data set. Three functions, `drop_cols`, `char_to_num`, and `convert_names`, have been included in the package to enable users to clean future editions of the VIUS data set in the same way that the data has been cleaned herein. Each function only takes one argument, the VIUS data set that the user would like to clean. Future releases
of the VIUS data set can be passed through these functions to prepare them for
use with the VIUSr package.

The data set `vius_raw_sample` has been included in the package to demonstrate
the data cleaning functions. This data set includes the first 100 rows of the
VIUS data set with no cleaning.

```{r example}
library(VIUSr)

head(vius_raw_sample)

vius_cleaned <- drop_cols(vius_raw_sample)
vius_cleaned <- char_to_num(vius_cleaned)
vius_cleaned <- convert_names(vius_cleaned)
head(vius_cleaned)
```


## VIUS Plotting

### Vehicle Body Type Averages
The `get_btype_graphs()` function computes weighted averages by vehicle body type and generates a horizontal bar chart.

```{r}
get_btype_graphs(vius1 = vius_cleaned,
                 var = `MPG`,
                 plot_title = "Average MPG by Body Type",
                 x_plot_label = "Body Type",
                 y_plot_label = "Average Miles Per Gallon")
```

### Mapping State-Level Data

The `get_state_maps()` function visualizes state-level VIUS data on an interactive U.S. map. This interactive plotting feature does not render inside the `github_document` format but please try it for yourself with the code below!

```r
get_state_maps(vius = vius_cleaned,
               var = `MPG`,
               var_label = "Fuel Volume",
               dollars = FALSE)
```


### Batch Plotting by State

To generate multiple bar plots by state, use the `get_state_graphs()` function. This is useful when you want to examine how a specific variable varies across different U.S. states.

Note: This function produces multiple plots and is best used in an interactive session or saved to files.

```r
Reference only
get_state_graphs(dataset = vius_data,
                  db_header = `MPG`,
                  states = c("CA", "TX", "NY"),
                  plot_title = "Average MPG by State",
                  x_plot_label = "Average MPG",
                  y_plot_label = "State")

```

### Package URL: https://github.com/jashonnew/VIUSr
